Install all dependencies if first WGEN simulation.

```{r}
# Set the working directory
#knitr::opts_knit$set(root.dir ="../")
setwd("..")
getwd()

#install all dependencies if first time running WGEN simulation
packages <- suppressWarnings(readLines("R_sessionInfo.txt"))
install.packages(packages)
```

# The next code chunk represents "process.meteorology.R"

This script provides the 'meteorology' input required for the WGEN run

Step 1: Importing raw meteorological time series

-   raw data files MUST BE formatted as: yyyy mm dd P[mm] Tmax[C] Tmin[C] \*.csv file

```{r}
# clear the current workspace
rm(list=ls())

# set working directory
setwd("..")
# print working directory for user
getwd()

# harcoded option
#dir.to.all.raw.files <- "./Data/raw.data.files/"

# select the folder using choose.dir() from utils package with custom title
dir.to.all.raw.files <- utils::choose.dir(caption = "Select folder containing raw data files")


# Check if a valid directory was selected
if (is.null(dir.to.all.raw.files)) {
  stop("No directory selected. Please select a valid directory.")
}

# Ensure the path ends with a backslash
if (substr(dir.to.all.raw.files, nchar(dir.to.all.raw.files), nchar(dir.to.all.raw.files)) != "\\") {
  dir.to.all.raw.files <- paste0(dir.to.all.raw.files, "\\")
}

# list of gridded location or station data
list.locations <- list.files(dir.to.all.raw.files, pattern = ".csv$") 
my.file <- read.table(paste0(dir.to.all.raw.files,list.locations[1]),sep= ",")
# reference the first .csv date
start_date <- paste(my.file[1,1],my.file[1,2],my.file[1,3],sep="-")
# reference the last .csv date
end_date <- paste(my.file[nrow(my.file),1],my.file[nrow(my.file),2],my.file[nrow(my.file),3],sep="-")

# create sequence of all dates in raw data
seq.of.dates <- seq(as.Date(start_date),as.Date(end_date),by = "day")
lst.grids.data <- list()


# create array of imported weather data
ascii.array.precip.temp <- array(NA,c(length(seq.of.dates),length(list.locations),3))
for (f in 1:length(list.locations)){

  my.file <- read.table(paste0(dir.to.all.raw.files,list.locations[f]),
                        sep= ",")
  ascii.array.precip.temp[,f,1] <- my.file[,4] # precip -- verify this
  ascii.array.precip.temp[,f,2] <- my.file[,5] # tmax -- verify this
  ascii.array.precip.temp[,f,3] <- my.file[,6] # tmin -- verify this
  prct.print <- round(f/length(list.locations)*100,2)
  print(paste('done -> location',f,'/',
              length(list.locations),', %',prct.print))
}

lst.import.datafile <- list('ascii.array.precip.temp'=ascii.array.precip.temp,
                            'seq.of.dates'=seq.of.dates,
                            'file.names'=list.locations)
# save weather data for future use
saveRDS(lst.import.datafile,
        file='./Data/processed.data.files/processed.meteorology/lst.import.datafile.rds')

#------------------------------------------------------------------ 
##/ Step 2: Organizing input data for WGEN runs \##

# import the inputted weather data
lst.import.datafile <- readRDS(file='./Data/processed.data.files/processed.meteorology/lst.import.datafile.rds')


# oragnize weather data for the basin processing
ascii.array.precip.temp <- lst.import.datafile$ascii.array.precip.temp
dates.weather <- lst.import.datafile$seq.of.dates
list.file.names <- lst.import.datafile$file.names
rm(lst.import.datafile) # for memory

years.weather <- as.numeric(format(format(dates.weather,'%Y')))
months.weather <- as.numeric(format(format(dates.weather,'%m')))
wateryears.weather <- (years.weather+1)*(months.weather>=10) + (years.weather)*(months.weather<10)

n.sites <- dim(ascii.array.precip.temp)[2]
prcp.site <- ascii.array.precip.temp[,,1]
tmax.site <- ascii.array.precip.temp[,,2]
tmin.site <- ascii.array.precip.temp[,,3]

rm(ascii.array.precip.temp)

prcp.basin <- apply(prcp.site,1,mean)
tmax.basin <- apply(tmax.site,1,mean)
tmin.basin <- apply(tmin.site,1,mean)


# create annual average for basin precipitation
prcp.basin.annual <- aggregate(prcp.basin,FUN=mean,by=list(years.weather),na.rm=T)
prcp.basin.annual[,2] <- scale(prcp.basin.annual[,2])[,1]

# create annual average for basin temperature mean.
tmin.basin.annual <- aggregate(tmin.basin,FUN=mean,by=list(years.weather),na.rm=T)
tmin.basin.annual[,2] <- scale(tmin.basin.annual[,2])[,1]
tmax.basin.annual <- aggregate(tmax.basin,FUN=mean,by=list(years.weather),na.rm=T)
tmax.basin.annual[,2] <- scale(tmax.basin.annual[,2])[,1]

save(list = ls(envir = environment(), all.names = TRUE), 
     file = "./Data/processed.data.files/processed.meteorology/processed.meteorology.RData",
     envir = environment())

gc()
```

# Next code chunk represents "config.simulations.R"

```{r}

# set working directory
setwd("..")

config.simulations <- function(){
  
  ######  --------------------------------------------------------------------------
  ######  --------------------------------------------------------------------------
  ######  ---------------> Simulation Length <------------------------------- ######
  ######  --------------------------------------------------------------------------
  ######  --------------------------------------------------------------------------
  {
    ##length of final simulated weather (in calendar years)##

    simulation.length <- suppressWarnings(read.table('SimulationLength.csv',header=TRUE,sep=","))

    number.years.long <- simulation.length$number_of_years_per_ensemble_member 
    # {e.g., 500, 1000, 2000, 3000, 5000 years,...} [note: current NHMM output (parametric) is for 1036 years; current non-parametric is for 3050 years]
    num.iter <- simulation.length$number_of_ensemble_members 
    # A single long trace (e.g., thousand years) is sufficient, although more can be developed if desired
  }
  
  
  ######  --------------------------------------------------------------------------
  ######  --------------------------------------------------------------------------
  ######---->Thermodynamic Climate Change Scenario<-------------------------######
  ######  --------------------------------------------------------------------------
  ######  --------------------------------------------------------------------------
  {
    ##-------------Define perturbations-------------##
    ## climate changes and jitter to apply:
    climate.change.scenarios <- suppressWarnings(read.table('ClimateChangeScenarios.csv',header=TRUE,sep=","))

    change.list <- data.frame("tc.max"=  climate.change.scenarios$max_temperature_change_degC, # {e.g., 0, 1, 2, ...} (changes in temperature)
                              "tc.min"=  climate.change.scenarios$min_temperature_change_degC, # {e.g., 0, 1, 2, ...} (changes in temperature)
                              "pmuc"= climate.change.scenarios$mean_precipitation_change_percent/100, # {e.g., 0, -.125, .125, ...} (changes in precipitation mean)
                              "pccc"= climate.change.scenarios$extreme_precipitation_scaling_rate_percent/100 # {e.g., 0, 0.07, 0.14, ...} (changes for precipitation extreme quantile -- CC)
    )
    ##----------------------------------------------##
  }
  
  
  ######  --------------------------------------------------------------------------
  ######  --------------------------------------------------------------------------
  ######  ---------------->Dates for Weather Data<--------------------------######
  ######  --------------------------------------------------------------------------
  ######  --------------------------------------------------------------------------
  {
    lst.import.datafile <- tryCatch(suppressMessages(readRDS(file='./Data/processed.data.files/processed.meteorology/lst.import.datafile.rds')),
    
                               
  error=function(e) {
    message('You have not yet run process.meteorology.R')
    print(e)
    })
    start.date.weather <- lst.import.datafile$seq.of.dates[1]
    end.date.weather <- lst.import.datafile$seq.of.dates[length(lst.import.datafile$seq.of.dates)]
  }
  
  ######  ----------------------------------------
  ######  ----------------------------------------
  ######----> Directories <---------######
  ######  ----------------------------------------
  ######  ----------------------------------------
  {
    dir.to.sim.files <- "./Data/simulated.data.files/WGEN.out"
    dir.create(file.path(dir.to.sim.files), showWarnings = FALSE)

    # directory to store output files
    dir.to.output.files <- './Data/output.data.files/'
    
    ##location of obs weather data (RData format): weather data (e.g., precip and temp) as matrices (time x lat|lon: t-by-number of grids); dates vector for time; basin average precip (see the example meteohydro file)

    path.to.processed.data.meteohydro <- "./Data/processed.data.files/processed.meteorology/processed.meteorology.RData"

  }
  
  
  ######  -----------------------------------------------------
  ######  -----------------------------------------------------
  ######  -------> WGEN Hyperparameters <--------------- ######
  ######  -----------------------------------------------------
  ######  -----------------------------------------------------
  {
    first.month <- 1
    last.month <- 12
    months <- seq(first.month,last.month) # Jan-Dec calendar year
    ##threshold for mixed Gamma-GPD population separation
    qq <- .99  
    
    #keep the jittering on
    to.jitter <- TRUE
    
    ##bootstrapping choices##
    window.size <- rep(3,length(months))   
    #the size of the window (in days) from which runs can be bootstrapped around the current day of simulation, by month: Jan -- Dec
    pr.trace <- 0.25     
    # {0.25 mm, 0.01 in} trace prcp threshold. 0.25 mm (for Livneh dataset); or 0.01 inches: lower threshold below which day is considered dry
    
    ##load in supporting functions
    files.sources = list.files("./Programs/functions",full.names = TRUE)
    my.functions <- sapply(files.sources, source)
  }
  
  ######  -------------------------------------------------------------------------
  ######  -------------------------------------------------------------------------
  ######----> Hyperparameters of the WRs Identification and Simulations<-----######
  ######  -------------------------------------------------------------------------
  ######  -------------------------------------------------------------------------
  {
    ##Choose below whether to use provided WRs (TRUE), or run WRs identification from scratch (FALSE)
    use.provided.WRs <- TRUE 
    #{TRUE, FALSE}: TRUE for the WRs already provided for the Pacific/North American sector in 1948-2021
    
    start.date.synoptic="1948-01-01"; end.date.synoptic="2021-12-31" 
    # from processed GPHA file


    start.date.WRs="1948-01-01"; end.date.WRs="2019-12-31" 
    # proper leap year orders (starting with leap year of 1948, ending a year before (i.e., 2019) the leap year of 2020)
    
    dates.WRs.specific <- seq(as.Date(start.date.WRs),as.Date(end.date.WRs),by="day")
    
    num.years.sim.WRs <- number.years.long # e.g., 500, 1000, 2000, 3000, 5000 years, etc [note: current NHMM output (parametric) is for 1036 years]
#change to user input?    
    
    dir.to.sim.WRs.files <- "./Data/simulated.data.files/WRs.out" # dir.to.sim.WRs.files
    
    num.iter.WRs <- num.iter   #number of iterations to simulate sequence of WRs
#change to user input? 
    path.to.processed.GPHAs <- './Data/processed.data.files/processed.hgt/hgt.500.Pacific.NorthAmer.synoptic.region_19480101_20211231.rds'
    
    # Covariates should be a matrix with the first column as dates, and the second column as 
    #      ... normalized pPC1 (scaled and centered)
    
    path.to.processed.SPI.PCs <- './Data/processed.data.files/processed.NHMM.data/paleo.norm.4.cold.PCs.dates_extracted.rds'
    
#define seasons and covariates for NHMM models of WRs
    cold.months <- c(11,12,1,2,3,4) # Nov-Apr
    warm.months <- c(5,6,7,8,9,10)  #May-Oct
    num.PCs <- 10
    seasons <- list(cold.months,warm.months)
    num_eofs.season <- rep(num.PCs,length(seasons))  #number of PCs to use for geopotential heights per season
    num_WRs.season <- c(7,3)    #number of WRs to fit per season
    
    ##Choose below whether through parametric or non-parametric way to create the simulated WRs ##
    dynamic.scenario  <- 0 # {0, 1}: 0: no dynamic change; 1: yes dynamic change 
    
    if (dynamic.scenario==0){
      ##===> Attempt #0 (thermodynamic only; no change to freq of WRs) ===##
      # #specify target change (as a percent) for WR probabilities
      WR_prob_change <- c(0,0,0,0,0,0,0,0,0,0) # between 0 and 1
      # #how close (in % points) do the WR frequencies (probabilities) need to be to the target
      lp.threshold <- 0.00001
      # #how much change do we allow in a sub-period sampling probability before incurring a larger penalty in the optimization
      piecewise_limit <- .02
      
      #   --------- NOTE: some of these hyper-parameters may need tuning depending on the dynamic climate change selected
      
      #   --------- Attempt with caution!!!!
    }else if(dynamic.scenario==1){
      ##===> Attempt #1 (dynamic scenario #1) ===##
      # #specify target change (as a percent) for WR probabilities (if, increasing WR3 in future)
      WR_prob_change <- c(0,0,.3,0,0,0,0,0,0,0) # between 0 and 1
      # #how close (in % points) do the WR frequencies (probabilities) need to be to the target
      lp.threshold <- 0.007
      # #how much change do we allow in a sub-period sampling probability before incurring a larger penalty in the optimization
      piecewise_limit <- .02
      
      ##===> Other option explored in final report===##
      # specify target change (as a percent) for WR probabilities (if, continuing their current trends in future)
      #WR_prob_change <- c(-0.09969436,  0.27467048,  0.33848792,
      #                    -0.28431861, -0.23549986,  0.03889970,
      #                    -0.05628958, 0.38059153, -0.16636739, -0.17995965) # between 0 and 1
      # how close (in % points) do the WR frequencies (probabilities) need to be to the target
      #lp.threshold <- 0.008
      # how much change do we allow in a sub-period sampling probability before incurring a larger penalty in the optimization
      #piecewise_limit <- .02
      
    }
    
    
    # returning the entire values inserted here to 'run.stochastic.weather.generator'
    values = as.list(environment())
    return(values)
  }
}
```

# Next code chunk represents "run.stochastic.weather.generator.R"

```{r}





# set working directory 
setwd("..")



# clear environment
rm(list=ls())

# import all dependecies which were installed by first code chunk
library(MASS) # Gamma fit
library(evmix) # GPD fit
library(eva) # GPD fit
library(depmixS4) # HMMs/NHMMs fit
library(markovchain) # HMMs/NHMMs fit
library(rebmix) # split/WRs
library(lpSolve) # lp optimization
library(mvtnorm) # MVN
library(lubridate) # dates
library(tictoc) # run time
library(moments) # computation
library(abind) # computation
#--------------------------------
library(zoo) # (plot)
library(fExtremes) # (plot)
library(scales) # (plot)
library(parallel) # (plot)
library(proxy) # (plot)
library(POT) # (plot) event-based computations
library(extRemes) # (plot)
library(ismev) # (plot)
library(readxl) # output



source("./Programs/config.simulations.R") # config file
lst <- config.simulations() # call in configuration inputs
for (i in 1:length(lst)) {assign(names(lst[i]), lst[[i]]) }; rm(lst)

rm(lst)

#--- Weather Regimes Module ---#
#use provided WRs
if (use.provided.WRs){

  final.NHMM.output <- readRDS('./Data/simulated.data.files/WRs.out/final.NHMM.non_param.output.rds')
  weather.state.assignments <- final.NHMM.output$WR.historical # this is the historical WRs 
  num.states <- length(unique(as.vector(weather.state.assignments)))    #number of WRs in the model
  dates.sim <- final.NHMM.output$dates.sim
  markov.chain.sim <- final.NHMM.output$WR.simulation
  dates.synoptics <- final.NHMM.output$dates.historical
#simulate your own WRs
} else{
  final.NHMM.output <- execute.WRs.non_param.NHMM()
  weather.state.assignments <- final.NHMM.output$WR.historical # this is the historical WRs 
  num.states <- length(unique(as.vector(weather.state.assignments)))    #number of WRs in the model
  dates.sim <- final.NHMM.output$dates.sim
  markov.chain.sim <- final.NHMM.output$WR.simulation
  dates.synoptics <- final.NHMM.output$dates.historical
}
rm(final.NHMM.output) # for memory


#--- Weather Generation Module ---#
suppressWarnings(execute.simulations())
# done. #


# EXTRA #
### Below are auxiliary functions to do a list of tasks
# - create sample figures for selected scenario
# - generate individual output files in tab or text delimited formats

#this is the scenario (i.e., the row in ClimateChangeScenarios.csv) for which to make plots and write out the data as .csv files
selected_scenario = 1

#--- figures ---#
# arguments are labels for x and y-axes
start_time <- Sys.time()
create.figures.baselines.stacked(scenario = selected_scenario)
Sys.time() - start_time


#--- outputs ---#
# YYYY, MM, DD, P(mm), Tmax(C), Tmin(C) in .csv individual lat/lon file #
# for simulated data #
create.delimited.outputs(scenario = selected_scenario)


```
